\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-07-07 eg6-letters/NIM/EG6_RTPC_NIM.tex
Add Stepan's comment

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nuclei) necessitates an upgrade of the photon detection system. 
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 
experiment in 2005~\cite{FX}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber (RTPC) was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in the RTPC. The CLAS detector supplemented with both IC and RTPC was used 
during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 300 MeV/c (12 MeV kinetic energy). Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fit inside the 230~mm diameter bore of the 
5 Tesla solenoid magnet.

@@

The new CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electric field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of cylindrical gas electron multipliers (GEM) and 
detected by the readout system 
on the external shell of the detector as illustrated in the 
Fig.~\ref{fig:RTPC_1_4}. The RTPC is segmented into two halves with 
independent GEM amplification systems\footnote{The foils were produced by 
Tech-Etch, Inc.} that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam axis; it 
      is a 284~mm long, 6~mm diameter straw with a 27-$\mu$m Kapton wall.
   \item The first gas gap covers the radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at $- 4260$~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low spark rate~\cite{GEM_ref}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 300~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Fig.~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

@@

The RTPC electron collection system had 3200 readout pads. These elements were
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} illustrates the configuration of the 5 by 4.45 mm pads,
where the shift between the rows was implemented to reduce aliasing. Each half of the 
RTPC had 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads were grouped to 16 channels pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driving the 6-m long ribbon 
cable that connects to it.

@@

Finally, during data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to to the trigger time, the resolution is equivalent to the ALTRO 
sampling time, 100 ns.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculated them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We this calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

@@

The second noise signature was a coherent noise affecting about 25\% of the 
amplifiers boards, when simultaneous hits in most of the 16 channels of the
board were recorded. An event-based technique to identify and remove this noise was developed based 
on counting simultaneous hits in each pre-amplifier group, and, if sufficiently large, 
perform a dynamic pedestal subtraction based on the average ADC of neighboring 
channels within this group.

>>>>

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nuclei) necessitates an upgrade of the photon detection system. 
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 
experiment in 2005~\cite{FX}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid magnet was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber (RTPC) was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in the RTPC. The CLAS detector supplemented with both IC and RTPC was used 
during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 300 MeV/c (12 MeV kinetic energy). Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fit inside the 230~mm diameter shell and cryostat 
wall of the solenoid magnet bore.

@@

The new CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electric field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of cylindrical gas electron multipliers (GEM) and 
detected by the readout system 
on the external shell of the detector as illustrated in the 
Fig.~\ref{fig:RTPC_1_4}. The RTPC is segmented into two halves with 
independent GEM amplification systems that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam axis; it 
      is a 284~mm long, 6~mm diameter Kapton straw with a 27-$\mu$m wall.
   \item The first gas gap covers the radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at $- 4260$~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low spark rate~\cite{GEM_ref}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 300~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers\footnote{The GEM foils were produced by 
Tech-Etch, Inc.}. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Fig.~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

@@

The RTPC electron collection system had 3200 readout pads. These elements were
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} illustrates the configuration of the 5 by 4.45 mm pads,
where the shift between the rows was implemented to reduce aliasing. Each half of the 
RTPC had 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads were grouped to 16 channels pre-amplifier boards. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driving the 6-m long ribbon 
cable that connects to it.

@@

Finally, during data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to the trigger time, the resolution is equivalent to the ALTRO 
sampling time, 100 ns.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculated them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used this calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometrical accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

@@

The second noise signature was a coherent noise affecting about 25\% of the 
pre-amplifiers boards, when simultaneous hits in most of the 16 channels of the
board were recorded. An event-based technique to identify and remove this noise was developed based 
on counting simultaneous hits in each pre-amplifier group, and, if sufficiently large, 
perform a dynamic pedestal subtraction based on the average ADC of neighboring 
channels within this group.
